\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\pgfplotsset{compat=1.17}
\setlist[itemize]{leftmargin=*}
\setlist[enumerate]{leftmargin=*}

\title{Inspectable Symbolic Compression for Long-Context Reasoning:\\A Mixed-Results Study}
\author{Anonymous Authors}
\date{\today}

\newcommand{\ism}{\textsc{ISM}}
\newcommand{\ar}{\textsc{AR}}
\newcommand{\cratio}{\textsc{CR}}
\newcommand{\es}{\textsc{ES}}
\newcommand{\full}{\textsc{Full}}
\newcommand{\summary}{\textsc{Summary}}
\newcommand{\keyword}{\textsc{Keyword}}

\begin{document}
\maketitle

\begin{abstract}
Large language models can answer questions over long documents, but long inputs
increase latency and token cost.  We study \emph{Inspectable Symbolic
Compression} (\ism), a prompt-only method that asks an instruction-tuned model
to rewrite a document as discrete symbols, dictionary definitions, and symbolic
relations.  The motivating hypothesis was that this format could preserve
task-relevant information while exposing units that can be removed, corrupted,
or swapped.  We report a mixed result.  In a controlled Synthetic Rule-QA
setting, the internal structure of \ism{} is used: flipping dictionary
conclusions and removing symbolic structure both significantly reduce accuracy.
However, when \ism{} is compared with a strong natural-language model-summary
baseline under the same token budgets, the summary is significantly more
accurate and more token-efficient.  In a reuse analysis, both \ism{} and
summaries become cheaper than full context from two questions onward, but the
summary dominates \ism{} in both cost and accuracy.  We therefore do not claim
that prompt-only \ism{} is a better compressor than natural-language summaries.
Instead, the study identifies a boundary: if symbolic compression is generated
by an LLM and then read again by an LLM, it remains in the same class as text
summarization.  The results point toward future work on executable semantic IRs
that are parsed, checked, transformed, and executed by programs rather than
only read by another LLM.
\end{abstract}

\section{Summary of Findings}

This paper is intentionally written as a mixed-results study.  The central
question is not whether \ism{} can be made to look good after the fact, but
which parts of the original hypothesis survive comparison against a strong
summary baseline.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.16\linewidth}p{0.48\linewidth}p{0.24\linewidth}}
\toprule
Question & Result & Interpretation \\
\midrule
RQ1 & \ism{} structure is used.  At dev scale, $\Delta_{\mathrm{map}}^{\mathrm{flip}}=+0.079$ ($p=0.032$) and $\Delta_{\mathrm{symbol}}=+0.108$ ($p=5\times10^{-4}$). & Positive, but dev-scale. \\
RQ3 & Model Summary beats \ism{} at all fixed budgets: paired summary--\ism{} gaps of $+0.250$, $+0.3125$, and $+0.375$ for 128, 256, and 512 tokens. & Negative for efficiency. \\
RQ4 & Reuse helps all cached representations, but Model Summary dominates \ism{} in both cost and accuracy. & Negative for \ism{}-specific reuse. \\
RQ2 & Dictionary swap with unseen labels remains unrun. & Future work. \\
\bottomrule
\end{tabular}
\caption{High-level status of the study.  The prompt-only \ism{} format is
functionally used, but it does not beat natural-language summaries as an
efficient compressor.}
\label{tab:summary}
\end{table}

\section{Introduction}

Long-context language models make it possible to answer questions over papers,
reports, and conversation histories, but the cost of repeatedly processing long
inputs remains high.  Prompt-compression methods reduce this cost by selecting
tokens, rewriting text, or learning compact continuous memories
\citep{jiang2023llmlingua,jiang2023longllmlingua,pan2024llmlingua2,li2023selective,mu2023gist,chevalier2023autocompressors,ge2024icae}.
These approaches are usually evaluated by two axes: how many tokens they save
and how much downstream accuracy they retain.

This leaves a different question underexplored.  If a compressed prompt
preserves enough information to answer correctly, can we identify which units
of meaning were preserved?  Can we damage or repair one unit and observe a
predictable change in the output?  Concept-bottleneck models show the value of
intervenable intermediate representations in supervised settings
\citep{koh2020concept,espinosa2022concept}; prompt compression has not usually
been evaluated through the same lens.

We therefore study \ism{}, which rewrites a document into three explicit parts:
symbols, dictionary entries, and relations.  A typical representation is:

\begin{quote}\small
\texttt{Z1 := if marker\_a=high and marker\_b=low then risk=HIGH}\\
\texttt{Z2 := if marker\_c=present then risk=LOW}\\
\texttt{RELATIONS: Z1 overrides Z2}
\end{quote}

The initial hope was that this representation might be both compact and
intervenable.  The experiments support only half of that hope.  The structure
is used by the model, but natural-language summaries are a stronger
prompt-only compressor.  This negative result is scientifically useful: it
clarifies that merely forcing an LLM to emit a symbolic-looking text format
does not create a qualitatively new computational object.  To become different
from a summary, symbolic compression likely has to become a program-readable
semantic IR.

\section{Related Work}

\paragraph{Prompt compression.}
LLMLingua and LongLLMLingua compress prompts by selecting or removing less
important tokens while attempting to preserve task-relevant information
\citep{jiang2023llmlingua,jiang2023longllmlingua}.  LLMLingua-2 distills
compression knowledge into a token-classification model and targets faithful,
task-agnostic compression \citep{pan2024llmlingua2}.  Selective Context
similarly removes low-information units to reduce inference cost
\citep{li2023selective}.  These methods motivate our fixed-budget comparison
against extractive and model-summary baselines.

\paragraph{Continuous context compression.}
Gist tokens, AutoCompressors, and ICAE compress prompts or contexts into
learned memory tokens or vectors \citep{mu2023gist,chevalier2023autocompressors,ge2024icae}.
Such representations can be highly compact but do not expose a natural unit
such as ``rule 3'' that can be deleted or counterfactually edited.  \ism{}
takes the opposite tradeoff: it uses ordinary text tokens and gives up some
density in exchange for explicit symbolic entries.

\paragraph{Intervenable representations.}
Concept Bottleneck Models and Concept Embedding Models insert human-meaningful
intermediate variables between inputs and predictions
\citep{koh2020concept,espinosa2022concept}.  \ism{} is related in spirit, but
the setting differs: the intermediate representation is generated from long
text at inference time and is evaluated as a prompt-compression artifact, not
as a supervised concept layer.

\section{Inspectable Symbolic Compression}

Given a document $x$, question $q$, and answer $y$, a compressor $C$ produces a
bounded representation $z=C(x;B)$ under token budget $B$, and a reasoner $R$
answers $\hat{y}=R(z,q)$.  For \ism{}, the representation is
\[
z_{\mathrm{ISM}} = (S,D,G),
\]
where $S=\{Z_1,\dots,Z_k\}$ is a set of symbol labels, $D$ maps labels to short
definitions, and $G$ stores relations such as precedence, exception, or
application order.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node/.style={draw, rounded corners, align=center, minimum height=1.0cm, font=\small},
  arrow/.style={->, thick}
]
\node[node, fill=blue!8] (doc) at (0,0) {Long\\document};
\node[node, fill=blue!8] (comp) at (3,0) {LLM\\compressor};
\node[node, fill=green!8] (ism) at (6.3,0) {Symbols\\Dictionary\\Relations};
\node[node, fill=orange!10] (reason) at (9.8,0) {LLM\\reasoner};
\node[node, fill=red!8] (interv) at (6.3,-1.8) {flip / blank\\derange / symbol-only};
\draw[arrow] (doc) -- (comp);
\draw[arrow] (comp) -- (ism);
\draw[arrow] (ism) -- (reason);
\draw[arrow] (ism) -- (interv);
\draw[arrow] (interv) -- (reason);
\end{tikzpicture}
\caption{\ism{} in the prompt-only setting.  The same LLM family generates the
compressed text and later reads it for question answering.  This design makes
the format inspectable, but it also places \ism{} in direct competition with
ordinary natural-language summaries.}
\label{fig:pipeline}
\end{figure}

We use several interventions.  \emph{Derangement} permutes dictionary entries
across labels and tests surface label binding.  \emph{Conclusion flip} changes
the conclusion of dictionary rules while preserving their trigger conditions,
and is the primary test of semantic dictionary dependence.  \emph{Blanking}
removes conclusion-bearing content.  \emph{Symbol-only} removes the dictionary
while preserving the symbolic relation string.

\section{Experimental Setup}

\paragraph{Data.}
Synthetic Rule-QA is a controlled long-document QA task in which each document
is generated from an underlying rule graph.  This graph lets us audit whether a
compressed representation preserves rule conditions, conclusions, and
relations.  The long-context profile uses documents of roughly 700--2000
tokens, making budgets of 64--512 tokens meaningful.  QASPER
\citep{dasigi2021qasper} is discussed as an external-validity target, but the
current paper reports only Synthetic Rule-QA experiments.

\paragraph{Model and decoding.}
The main compressor and reasoner are Qwen2.5-7B-Instruct in 4-bit inference,
with greedy decoding.  Compression outputs are regenerated rather than
truncated when they exceed the budget.  We report accuracy, accuracy retention
(\ar), compression ratio (\cratio), and efficiency score $\es=\ar/\cratio$.
Paired comparisons use bootstrap confidence intervals and McNemar tests
\citep{mcnemar1947note,efron1994introduction}.

\paragraph{Baselines.}
The central baseline is Model Summary: the same LLM produces a natural-language
summary under the same budget.  We also include keyword extraction and an
oracle gold summary derived from the hidden rule graph.  LLMLingua-2 is part of
the intended comparison suite but is not run in the reported dev-scale results.

\section{Results}

\subsection{RQ1: The Symbolic Structure Is Functionally Used}

The first ablation study asks whether the model actually uses the dictionary
and symbolic structure.  Early null results were misleading: the compressor
often dropped conclusions, and a label-derangement corruption preserved the
set of rule definitions.  After fixing the compressor and replacing
derangement with semantic conclusion flips, the signal became clear.

\begin{table}[t]
\centering
\small
\begin{tabular}{lccc}
\toprule
Condition & Accuracy & \ar & \cratio \\
\midrule
Full Context & 0.750 & 1.000 & 1.000 \\
Full Symbol + Dict & 0.446 & 0.594 & 0.745 \\
Deranged Dict & 0.375 & 0.500 & 0.745 \\
Flipped Dict & 0.367 & 0.489 & 0.745 \\
Blank Dict & 0.250 & 0.333 & 0.215 \\
Symbol Only & 0.413 & 0.550 & 0.079 \\
Random Symbol & 0.304 & 0.406 & 0.738 \\
\bottomrule
\end{tabular}
\caption{Dictionary ablation, dev scale-up: 120 documents, 240 questions, and
1680 predictions.  Evidence is stored under
\texttt{docs/evidence/ablation-qwen7b-N120/}.}
\label{tab:ablation}
\end{table}

At $N=240$ paired questions, flipping dictionary conclusions lowers accuracy
relative to the normal dictionary
($\Delta_{\mathrm{map}}^{\mathrm{flip}}=+0.079$, 95\% CI $[0.013,0.146]$,
$p=0.032$).  Symbol-only representations also beat length-matched random
symbols ($\Delta_{\mathrm{symbol}}=+0.108$, 95\% CI $[0.050,0.167]$,
$p=5\times10^{-4}$).  In contrast, derangement remains non-significant
($p=0.159$), consistent with the diagnosis that it tests surface label binding
rather than semantic content.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.92\linewidth,
  height=5cm,
  ymin=-0.05,
  ymax=0.18,
  ybar,
  bar width=16pt,
  symbolic x coords={derange,flip,symbol},
  xtick=data,
  ylabel={Accuracy difference},
  nodes near coords,
  nodes near coords align={vertical},
]
\addplot coordinates {(derange,0.071) (flip,0.079) (symbol,0.108)};
\end{axis}
\end{tikzpicture}
\caption{Key RQ1 contrasts.  The semantic flip and symbol-vs-random contrasts
are positive; derangement is treated as a secondary label-binding diagnostic.}
\label{fig:rq1}
\end{figure}

\subsection{RQ3: Natural-Language Summaries Beat \ism{} Under Fixed Budgets}

The fixed-budget experiment compares \ism{}, Model Summary, keyword extraction,
and oracle summaries at budgets of 128, 256, and 512 tokens.  Budget 64 is
excluded from the primary comparison because \ism{} has a floor of roughly 100
tokens and fails to fit.

\begin{table}[t]
\centering
\small
\begin{tabular}{lccc}
\toprule
Method & 128 & 256 & 512 \\
\midrule
Oracle Gold Summary & 0.732 / 11.0 & 0.732 / 11.0 & 0.732 / 11.0 \\
Model Summary & 1.036 / 51.3 & 1.125 / 17.3 & 1.196 / 13.3 \\
Keyword Extract & 0.714 / 7.4 & 0.732 / 4.2 & 0.750 / 3.6 \\
\ism{} & 0.679 / 9.4 & 0.679 / 9.3 & 0.661 / 8.9 \\
\bottomrule
\end{tabular}
\caption{Fixed-budget dev pilot.  Each cell is \ar{} / \es{} relative to full
context.  The primary analysis is the paired summary-vs-\ism{} contrast in
Table~\ref{tab:paired}.}
\label{tab:fixed}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{ccccc}
\toprule
Budget & $\Delta=\mathrm{Acc}_{summary}-\mathrm{Acc}_{ISM}$ & 95\% CI & McNemar $p$ & Discordant \\
\midrule
128 & +0.2500 & [0.100, 0.400] & 0.0029 & 31 / 11 \\
256 & +0.3125 & [0.175, 0.450] & $1.1\times10^{-4}$ & 33 / 8 \\
512 & +0.3750 & [0.250, 0.500] & $2.3\times10^{-7}$ & 33 / 3 \\
\bottomrule
\end{tabular}
\caption{Primary RQ3 paired contrast on the same 80 questions.  Model Summary
significantly outperforms \ism{} at every budget.  Evidence is stored under
\texttt{docs/evidence/fixed-budget-N40/}.}
\label{tab:paired}
\end{table}

This result is not merely an artifact of using full-context accuracy as the
denominator for \ar.  A filler sanity check showed that full-context accuracy
can degrade as neutral filler grows, but the primary fixed-budget result
compares summary and \ism{} on the same questions after both have removed
filler.  Model Summary still wins.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.92\linewidth,
  height=5cm,
  ymin=0,
  ymax=0.45,
  xlabel={Budget},
  ylabel={Summary accuracy advantage},
  xtick={128,256,512},
  grid=major,
]
\addplot+[mark=*] coordinates {(128,0.25) (256,0.3125) (512,0.375)};
\end{axis}
\end{tikzpicture}
\caption{Paired Model Summary minus \ism{} accuracy difference.  The advantage
increases with budget in the dev pilot.}
\label{fig:summarywins}
\end{figure}

\subsection{RQ4: Reuse Helps Cached Text, but Not \ism{} Specifically}

Reuse is analytically evaluated by charging one compression pass per document
and then serving multiple questions from the cached representation.  This helps
all cached representations.  However, Model Summary again dominates \ism{}.

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrrrr}
\toprule
Method & $n=1$ & $n=8$ & $n=32$ & $n=64$ & Acc. \\
\midrule
Full Context & 1354 & 10828 & 43312 & 86624 & 0.700 \\
Model Summary & 1526 & 2168 & 4369 & 7305 & 0.787 \\
\ism{} & 1548 & 2265 & 4725 & 8005 & 0.475 \\
Oracle Gold & 1530 & 2184 & 4428 & 7420 & 0.512 \\
Keyword Extract & 1820 & 3492 & 9225 & 16868 & 0.512 \\
\bottomrule
\end{tabular}
\caption{Reuse cost-accuracy tradeoff at budget 256.  All cached methods beat
full context from two questions onward, but Model Summary is cheaper and more
accurate than \ism{}.}
\label{tab:reuse}
\end{table}

\section{Discussion}

\paragraph{What worked.}
The \ism{} format is not inert.  Once the compressor is forced to preserve
conclusions and the corruption tests semantic content, both the dictionary and
symbolic structure show functional effects.  The early null result was a
measurement failure, not a clean refutation of the idea that the representation
can be read by the model.

\paragraph{What did not work.}
The stronger claim, that prompt-only \ism{} is a better compressor than natural
language summaries, is not supported.  In this setting, both \ism{} and Model
Summary are texts generated by an LLM and read by an LLM.  Within that class,
ordinary summaries are shorter, more accurate, and cheaper to reuse.

\paragraph{What this implies.}
If symbolic compression is to become more than a verbose summary format, it
must leave the ``LLM writes text, LLM reads text'' loop.  A future semantic IR
should be parsed into an AST, checked for invariants, transformed through
local programmatic edits, and executed by a symbolic engine when possible.  The
relevant contrast would then no longer be ``Can the LLM read this prompt?'' but
``Can the LLM compile the document into an executable representation?''

\section{Limitations}

The main experiments are dev-scale and use a synthetic rule domain rather than
a full registered-scale run.  QASPER is specified but not evaluated in the
reported results.  LLMLingua-2 is planned but absent from the dev pilot.
Several raw GPU artifacts are retained by hash and reproducibility commands
rather than fully stored in the repository.  Finally, the study evaluates a
prompt-only \ism{} format; it does not test a parser/checker/executor system,
which is precisely the future direction suggested by the negative result.

\section{Conclusion}

We studied Inspectable Symbolic Compression as a candidate long-context
compression format.  The result is mixed.  \ism{} representations are
functionally used: damaging semantic dictionary content and removing symbolic
structure both hurt performance.  But \ism{} is not a better prompt-only
compressor than a natural-language model summary.  Summaries win in fixed
budget accuracy, efficiency score, and reuse cost-accuracy frontier.  The
honest conclusion is that prompt-only symbolic compression fails as an
efficiency story, while pointing to a sharper future direction: executable
semantic IRs that are checked, edited, and run by programs outside the LLM.

\section*{Reproducibility}

The repository stores evidence under \texttt{docs/evidence/}.  The most
important runs are \texttt{ablation-qwen7b-N120/}, \texttt{fixed-budget-N40/},
\texttt{reuse-N40/}, and the diagnostic report
\texttt{docs/reviews/llm-ism-diagnostic.md}.  The LaTeX and PDF artifacts for
this paper are stored under \texttt{paper/}.

\bibliographystyle{plainnat}
\bibliography{references}

\appendix
\section{Pre-registered Criteria and Dev-Scale Status}

The original criteria were: (1) \ism{} should outperform Model Summary in
\ar{} or \es; (2) Symbol Only should beat Random Symbol and Full Symbol + Dict
should beat corrupted dictionary conditions; and (3) unseen swapped labels with
a dictionary should retain most original-label performance.  Criterion (1)
fails in the fixed-budget and reuse experiments.  Criterion (2) is met under
the amended semantic-flip corruption.  Criterion (3) remains untested.

\section{Future Semantic-IR Experiments}

Two follow-up experiments would move beyond the prompt-only setting.
First, a compile-and-execute experiment would parse LLM-generated \ism{} into an
AST and answer Synthetic Rule-QA questions with a symbolic executor.  Second, a
programmatic-editability experiment would corrupt one rule, repair it through a
deterministic rule-id edit, validate the schema, and measure targeted recovery
and collateral damage.  These tests would compare machine-checkable symbolic
IRs against free-form summaries without falsely claiming that summaries cannot
be manually edited.

\end{document}

